### 2025-07-22 getting-started.qmd
20250721 - stacking correlation matrix

   install.packages(c(
     "petersenlab","remotes","knitr","rmarkdown","tidyverse","nflverse",
     "tidymodels","easystats","broom","broom.mixed","psych","downlit","xml2",
     "gsisdecoder","progressr","DescTools","pwr","pwrss","WebPower","XICOR",
     "dagitty","ggdag","ggtext","gghighlight","ggExtra","grid","patchwork",
     "pROC", "lme4","lmerTest","MuMIn","emmeans","pbkrtest","sjstats",
     "AICcmodavg","rstan","brms","tidybayes","bbmle","fitdistrplus","sn",
     "mclust","magrittr","viridis","viridisLite","msir","plotly","webshot2",
     "quantmod","fPortfolio","NMOF","nFactors","xts","zoo","forecast","stringi",
     "parallelly","doParallel","missRanger","ggridges","powerjoin","caret",
     "LongituRF","gpboost","corrplot","mgcv","rms","car","lavaan","lavaanPlot",
     "lavaangui","mice","miceadds","interactions","robustbase","ordinal","MASS",
     "data.table","future","future.apply","SimDesign","domir"))
   ```
1. Some necessary packages, including the [`ffanalytics` package](https://github.com/FantasyFootballAnalytics/ffanalytics) [@R-ffanalytics], are hosted in GitHub (and are not hosted on the Comprehensive R Archive Network [CRAN]) and thus need to be installed using the following code (after installing the `remotes` package [@R-remotes] above)[^petersenlabPackageGitHub]:
   ```{r}
   #| eval: false

>>>>

   install.packages(c(
     "petersenlab","remotes","knitr","rmarkdown","tidyverse","nflverse",
     "tidymodels","easystats","broom","broom.mixed","psych","downlit","xml2",
     "gsisdecoder","progressr","DescTools","pwr","pwrss","WebPower","XICOR",
     "dagitty","ggdag","ggtext","gghighlight","ggExtra","grid","patchwork",
     "pROC", "lme4","lmerTest","MuMIn","emmeans","pbkrtest","sjstats",
     "AICcmodavg","rstan","brms","tidybayes","bbmle","fitdistrplus","sn",
     "mclust","magrittr","viridis","viridisLite","msir","plotly","webshot2",
     "quantmod","fPortfolio","NMOF","nFactors","xts","zoo","forecast","stringi",
     "parallelly","doParallel","missRanger","ggridges","powerjoin","caret",
     "LongituRF","gpboost","corrplot","mgcv","rms","car","lavaan","lavaanPlot",
     "lavaangui","mice","miceadds","interactions","robustbase","ordinal","MASS",
     "data.table","future","future.apply","SimDesign","domir","GGally"))
   ```
1. Some necessary packages, including the [`ffanalytics` package](https://github.com/FantasyFootballAnalytics/ffanalytics) [@R-ffanalytics], are hosted in GitHub (and are not hosted on the Comprehensive R Archive Network [CRAN]) and thus need to be installed using the following code (after installing the `remotes` package [@R-remotes] above)[^petersenlabPackageGitHub]:
   ```{r}
   #| eval: false

### 2025-07-22 index.qmd
20250721 - stacking correlation matrix

# Automatically create a bib database for R packages
packagesForBibliography <- c(
  "petersenlab", "ffanalytics",
  "knitr", "rmarkdown", "broom", "broom.mixed", "psych", "downlit", "xml2", 
  "gsisdecoder", "progressr", "DescTools", "pwr", "pwrss", "WebPower", "XICOR", 
  "dagitty", "ggdag", "ggtext", "gghighlight", "ggExtra", "grid", "patchwork", 
  "pROC",  "lme4", "lmerTest", "MuMIn", "emmeans", "pbkrtest", "sjstats", 
  "AICcmodavg", "rstan", "brms", "tidybayes", "bbmle", "fitdistrplus", "sn", 
  "mclust", "magrittr", "viridis", "viridisLite", "msir", "plotly", "webshot2", 
  "quantmod", "fPortfolio", "NMOF", "nFactors", "xts", "zoo", "forecast",
  "stringi", "parallelly", "doParallel", "missRanger", "ggridges", "powerjoin", 
  "caret", "LongituRF", "gpboost", "cmdstanr", "corrplot", "mgcv", "rms", 
  "car", "lavaan", "lavaanPlot", "mice", "miceadds", "interactions", 
  "robustbase", "ordinal", "MASS", "data.table", "future", "future.apply", 
  "SimDesign", "domir",
  "tidyverse", "ggplot2", "dplyr", "lubridate", # packages that are part of the tidyverse ecosystem
  "nflverse", "nflfastR", "nflseedR", "nfl4th", "nflreadr", "nflplotR", # packages that are part of the nflverse ecosystem
  "tidymodels", "rsample", "parsnip", "recipes", "tune", "yardstick", # packages that are part of the tidymodels ecosysyem
  "easystats", "see", "performance", "parameters", "correlation", "effectsize", # packages that are part of the easystats ecosystem
  "learnr", "remotes", "shiny", "PearsonDS", "lavaangui", "dynamic" # packages that are cited but not used for computation (and thus do not need to be installed by me or readers)
)

>>>>

# Automatically create a bib database for R packages
packagesForBibliography <- c(
  "petersenlab", "ffanalytics",
  "knitr", "rmarkdown", "broom", "broom.mixed", "psych", "downlit", "xml2", 
  "gsisdecoder", "progressr", "DescTools", "pwr", "pwrss", "WebPower", "XICOR", 
  "dagitty", "ggdag", "ggtext", "gghighlight", "ggExtra", "grid", "patchwork", 
  "pROC",  "lme4", "lmerTest", "MuMIn", "emmeans", "pbkrtest", "sjstats", 
  "AICcmodavg", "rstan", "brms", "tidybayes", "bbmle", "fitdistrplus", "sn", 
  "mclust", "magrittr", "viridis", "viridisLite", "msir", "plotly", "webshot2", 
  "quantmod", "fPortfolio", "NMOF", "nFactors", "xts", "zoo", "forecast",
  "stringi", "parallelly", "doParallel", "missRanger", "ggridges", "powerjoin", 
  "caret", "LongituRF", "gpboost", "cmdstanr", "corrplot", "mgcv", "rms", 
  "car", "lavaan", "lavaanPlot", "mice", "miceadds", "interactions", 
  "robustbase", "ordinal", "MASS", "data.table", "future", "future.apply", 
  "SimDesign", "domir", "GGally",
  "tidyverse", "ggplot2", "dplyr", "lubridate", # packages that are part of the tidyverse ecosystem
  "nflverse", "nflfastR", "nflseedR", "nfl4th", "nflreadr", "nflplotR", # packages that are part of the nflverse ecosystem
  "tidymodels", "rsample", "parsnip", "recipes", "tune", "yardstick", # packages that are part of the tidymodels ecosysyem
  "easystats", "see", "performance", "parameters", "correlation", "effectsize", # packages that are part of the easystats ecosystem
  "learnr", "remotes", "shiny", "PearsonDS", "lavaangui", "dynamic" # packages that are cited but not used for computation (and thus do not need to be installed by me or readers)
)

### 2025-07-22 modern-portfolio-theory.qmd
20250721 - stacking correlation matrix

```{r}
library("quantmod")
library("fPortfolio")
library("NMOF")
library("tidyverse")
```

@@

load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/players_projectedPoints_seasonal.RData", fsep = ""))
load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/player_stats_weekly.RData", fsep = ""))
load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/nfl_playerIDs.RData", fsep = ""))
```

```{r}
load(file = "./data/players_projectedPoints_seasonal.RData")
load(file = "./data/player_stats_weekly.RData")
load(file = "./data/nfl_playerIDs.RData")
```

@@

However, sometimes having two players on the same team might be beneficial because some positions may be uncorrelated or even negatively correlated, which can also reduce risk.
For instance, the performance of the Tight End and Running Back on the same team tends to be slightly negatively correlated, so it might not be a bad idea to start the Tight End and Running Back from the same team.
For a correlation matrix of all positions on the team, see: <https://assets-global.website-files.com/5f1af76ed86d6771ad48324b/607a4434a565aa7763bd1312_AndyAsh-Sharpstack-RPpaper.pdf> [@Sherman2021; archived at <https://perma.cc/JQ6G-KSRT>] and <https://www.4for4.com/2018/preseason/definitive-guide-stacking-draftkings> [@4for4Staff2018; archived at <https://perma.cc/JZH3-ZM5V>].

@@

>>>>

```{r}
library("quantmod")
library("fPortfolio")
library("NMOF")
library("GGally")
library("tidyverse")
```

@@

load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/players_projectedPoints_seasonal.RData", fsep = ""))
load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/player_stats_seasonal.RData", fsep = ""))
load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/player_stats_weekly.RData", fsep = ""))
load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/nfl_actualFantasyPoints_weekly.RData", fsep = ""))
load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/nfl_playerIDs.RData", fsep = ""))
load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/nfl_schedules.RData", fsep = ""))
```

```{r}
load(file = "./data/players_projectedPoints_seasonal.RData")
load(file = "./data/player_stats_seasonal.RData")
load(file = "./data/player_stats_weekly.RData")
load(file = "./data/nfl_actualFantasyPoints_weekly.RData")
load(file = "./data/nfl_playerIDs.RData")
load(file = "./data/nfl_schedules.RData")
```

@@

However, sometimes having two players on the same team might be beneficial because some positions may be uncorrelated or even negatively correlated, which can also reduce risk.
For instance, the performance of the Tight End and Running Back on the same team tends to be slightly negatively correlated, so it might not be a bad idea to start the Tight End and Running Back from the same team.
For a correlation matrix of all positions on the team, see: <https://assets-global.website-files.com/5f1af76ed86d6771ad48324b/607a4434a565aa7763bd1312_AndyAsh-Sharpstack-RPpaper.pdf> [@Sherman2021; archived at <https://perma.cc/JQ6G-KSRT>] and <https://www.4for4.com/2018/preseason/definitive-guide-stacking-draftkings> [@4for4Staff2018; archived at <https://perma.cc/JZH3-ZM5V>].
We generate a correlation matrix of positions in @sec-stacking.

@@

### Stacking {#sec-stacking}

Stacking involves selecting players on the same team (e.g., multiple players on the Dallas Cowboys).
To help inform the process of stacking, we evaluate the inter-correlation of performance in a given week across players on the same team and on opposing teams.
We determine QB1, RB1, RB2, WR1, WR2, WR3, TE1, and K1 for each combination of team and season based on how many fantasy points a player scored over the entire season.

```{r}
# Identify the QB1, RB1, RB2, WR1, WR2, WR3, TE1, and K1 for each team-season combination
qb1ByTeamSeasonPosition <- player_stats_seasonal %>% 
  filter(!is.na(player_id) & !is.na(team) & !is.na(season) & !is.na(position_group) & !is.na(fantasyPoints)) %>% 
  filter(position_group %in% c("QB")) %>% 
  group_by(team, season, position_group) %>% 
  arrange(-fantasyPoints) %>% 
  slice_max(
    fantasyPoints,
    with_ties = FALSE) %>%
  ungroup() %>% 
  arrange(season, team, position_group) %>% 
  select(team, season, position_group, player_id) %>% 
  mutate(position = "QB1")

rbsByTeamSeasonPosition <- player_stats_seasonal %>% 
  filter(!is.na(player_id) & !is.na(team) & !is.na(season) & !is.na(position_group) & !is.na(fantasyPoints)) %>% 
  filter(position_group %in% c("RB")) %>% 
  group_by(team, season, position_group) %>% 
  arrange(-fantasyPoints) %>% 
  slice_max(
    fantasyPoints,
    n = 2,
    with_ties = FALSE) %>% 
  mutate(position = paste0("RB", row_number())) %>% 
  ungroup() %>% 
  arrange(season, team, position_group) %>% 
  select(team, season, position_group, player_id, position)

wrsByTeamSeasonPosition <- player_stats_seasonal %>% 
  filter(!is.na(player_id) & !is.na(team) & !is.na(season) & !is.na(position_group) & !is.na(fantasyPoints)) %>% 
  filter(position_group %in% c("WR")) %>% 
  group_by(team, season, position_group) %>% 
  arrange(-fantasyPoints) %>% 
  slice_max(
    fantasyPoints,
    n = 3,
    with_ties = FALSE) %>% 
  mutate(position = paste0("WR", row_number())) %>% 
  ungroup() %>% 
  arrange(season, team, position_group) %>% 
  select(team, season, position_group, player_id, position)

te1ByTeamSeasonPosition <- player_stats_seasonal %>% 
  filter(!is.na(player_id) & !is.na(team) & !is.na(season) & !is.na(position_group) & !is.na(fantasyPoints)) %>% 
  filter(position_group %in% c("TE")) %>% 
  group_by(team, season, position_group) %>% 
  arrange(-fantasyPoints) %>% 
  slice_max(
    fantasyPoints,
    with_ties = FALSE) %>%
  ungroup() %>% 
  arrange(season, team, position_group) %>% 
  select(team, season, position_group, player_id) %>% 
  mutate(position = "TE1")

k1ByTeamSeasonPosition <- player_stats_seasonal %>% 
  filter(!is.na(player_id) & !is.na(team) & !is.na(season) & !is.na(position_group) & !is.na(fantasyPoints)) %>% 
  filter(position_group %in% c("K")) %>% 
  group_by(team, season, position_group) %>% 
  arrange(-fantasyPoints) %>% 
  slice_max(
    fantasyPoints,
    with_ties = FALSE) %>%
  ungroup() %>% 
  arrange(season, team, position_group) %>% 
  select(team, season, position_group, player_id) %>% 
  mutate(position = "K1")

topRankedPlayersByTeamSeasonPosition <- dplyr::bind_rows(
  qb1ByTeamSeasonPosition,
  rbsByTeamSeasonPosition,
  wrsByTeamSeasonPosition,
  te1ByTeamSeasonPosition,
  k1ByTeamSeasonPosition
)

fantasyPointsForTopRankedPlayers_weekly <- dplyr::left_join(
  topRankedPlayersByTeamSeasonPosition,
  player_stats_weekly %>% select(team, season, week, player_id, fantasyPoints, game_id),
  by = c("team","season","player_id")
)

# add game_id to dst data

# first attempt: match on home_team
dst_with_game_home <- nfl_actualFantasyPoints_dst_weekly %>%
  left_join(
    nfl_schedules %>% select(game_id, season, week, home_team),
    by = c("season", "week", "team" = "home_team")
  )

# second attempt: match on away_team
dst_with_game_away <- nfl_actualFantasyPoints_dst_weekly %>%
  left_join(
    nfl_schedules %>% select(game_id, season, week, away_team),
    by = c("season", "week", "team" = "away_team")
  )

# Combine the two (one will have NA for game_id, the other will have the match)
dst_with_game_id <- dplyr::bind_rows(
  dst_with_game_home,
  dst_with_game_away) %>%
  filter(!is.na(game_id)) # keep only matched rows

dst_with_game_id_subset <- dst_with_game_id %>% 
  select(team, season, week, fantasyPoints, game_id) %>%
  mutate(
    position = "DST",
    position_group = "DST")

fantasyPointsSameTeam_weekly <- dplyr::bind_rows(
  fantasyPointsForTopRankedPlayers_weekly,
  dst_with_game_id_subset
) %>% 
  select(-player_id) %>% 
  filter(!is.na(game_id))

fantasyPointsOpponent_weekly <- fantasyPointsSameTeam_weekly %>%
  rename(
    opp_team = team,
    opp_fantasyPoints = fantasyPoints
  )

fantasyPointsCombined_weekly <- fantasyPointsSameTeam_weekly %>%
  inner_join(
    fantasyPointsOpponent_weekly %>% select(game_id, opp_team, position, opp_fantasyPoints),
    by = c("game_id", "position")
  ) %>%
  filter(team != opp_team) # exclude matching to own team

# Pivot to wide format for team positions
team_points_wide <- fantasyPointsCombined_weekly %>%
  select(team, season, week, position, fantasyPoints) %>%
  tidyr::pivot_wider(
    names_from = position,
    values_from = fantasyPoints)

# Pivot to wide format for opponent positions
opp_points_wide <- fantasyPointsCombined_weekly %>%
  select(team, season, week, position, opp_fantasyPoints) %>%
  tidyr::pivot_wider(
    names_from = position,
    values_from = opp_fantasyPoints,
    names_prefix = "opp_")

# Join both so each row has team and opponent positions side by side
combined_points_wide <- dplyr::left_join(
  team_points_wide,
  opp_points_wide,
  by = c("team", "season", "week")
)
```

Here is a correlation matrix of players' weekly performance across positions within a given team and their opponent:

```{r}
cor(
  combined_points_wide %>% select(QB1:opp_DST),
  use = "pairwise.complete.obs") %>% 
  round(., 2)
```

In @fig-correlationMatrixPlot, we depict a correlation matrix plot using the `ggcorr()` function of the `GGally` [@R-GGally] package.

```{r}
#| label: fig-correlationMatrixPlot
#| fig-cap: "Correlation Matrix Plot."
#| fig-alt: "Correlation Matrix Plot."

GGally::ggcorr(combined_points_wide %>% select(QB1:opp_DST))
```
